Analysing Constraint Grammars with a SAT-solver

### 2015-10-25 paper/lrec/eval_futurework.tex
more cleanup

We tested three grammars to find rules that cannot apply. The smallest
grammar was
Dutch\footnote{\url{https://svn.code.sf.net/p/apertium/svn/languages/apertium-nld/apertium-nld.nld.rlx}},
with 59 rules; second was
Spanish\footnote{\url{https://svn.code.sf.net/p/apertium/svn/languages/apertium-spa/apertium-spa.spa.rlx}}with
279 rules, and the largest was
Finnish\footnote{\url{https://github.com/flammie/omorfi/blob/master/src/vislcg3/omorfi.cg3}},
with 1185 rules. We left out \textsc{add}, \textsc{map} and other rule types introduced in CG-3, and only tested the traditional \textsc{remove} and \textsc{select} rules.

\begin{table}[]
\centering
\begin{tabular}{|l|l|l|l|}
\hline
                      & \textsc{nld}  & \textsc{spa}  & \textsc{fin}  \\ \hline
\# rules      & 59              & 279      & 1185     \\ \hline
\# readings      & 276             & 2191     & 5719    \\ \hline
\# conflicts       & 1               & 1        & 22    \\ \hline
\clock{} all rules       & 7s              & 3min 22s    & 3h 26min    \\ \hline
\clock{} last rule       & 4s              & 7s                & 3min 40s    \\ \hline
% This is with just 2000ish variables for Finnish, below is with 5700

@@

\paragraph{Correctness} We tested all three grammars if they contain
rules that cannot be applied, as described in the previous section.
This test revealed problems in all grammars, and for the smaller grammars, we were able to verify manually that the results are true.

@@

The program found 22 inapplicable rules in the Finnish grammar, but we
didn't have time to investigate all of them.
We could verify a number of tagset errors---the grammar was originally written in 1995, and converted to the Omorfi tagset by \cite{pirinen2015}. 
For example, the Omorfi tagset marks a possessive suffix with \texttt{Px}, e.g. \texttt{PxPl2} for second person plural, but the grammar used the old convention \texttt{PL2} in the rules.
As a result, all rules with possessive suffixes were detected as impossible.
In addition, we suspect that some of the problems come from our naive
implementation of rules with * in its context; creating the same
symbolic sentence for condition \texttt{1*} and \texttt{1}.
%For a condition situated $n$ or more words to the left or right, we only create a word exactly $n$ words away. This is something to be fixed in the future.

\paragraph{Running time} Currently the speed of the check is an obvious issue,
especially for the Finnish grammar. 
The 3 hours running time is achieved with using 2100 readings for the
symbolic words; instead of the lexicon, we only used the readings
specified in the grammar: for example, the rule \texttt{REMOVE ("bear"
  noun sg) IF (-1 ())
Using all 6000 readings from the lexicon, the program
ran the analysis for the last rule in 30 minutes, making the running
time for 1000+ rules potentially days.

@@

The SAT problem for checking one rule grows depending on how many rules come before it, how long the context is, and how many tag combinations are there. The running time of the last rule gives some indication ...

Not all questions are so expensive.
For instance, checking the tag sets in the Finnish grammar doesn't require the context of the other rules at all,
 The majority of the correctly detected errors in the Finnish grammar were rule-internal---these errors don't need the context of the other rules at all.

%All rules is the total time to check every rule against the previous rules, and last rule % shows how long it took to check the last rule.
% As indicated by the total time and the time to check the last rule, the size of the SAT problem grows according to how far in the sequence the rule is.

% As seen in table~\ref{table:res}, checking every rule in the Finnish grammar takes currently 3.5 hours.
% A rule takes longer to run if it appears later in the rule sequence: 1000 previous rules is a larger SAT problem than 10 previous rules.
% This can be seen from the time it took to compute the result for the last rule. For Finnish, the last rule took almost 4 minutes, but if we divide the time with the number of rules, the average time for one rule is just 10 seconds.
% We also measured the running times for the longest rules in the grammar.
%\todo{find longest rule, change place of that rule, see what matters more. Promise: we can be smarter about SAT things.}

\paragraph{Future work} We will continue improving the performance,
and make sure that the implementation for the basic rules is correct;
for instance, fixing the earlier mentioned issue with a non-exact
position.
We will also improve usability
After this, we want to handle the full expressivity of CG-3, with \textsc{map}, \textsc{add} and \textsc{substitute} rules.
At that stage, it would be interesting to test our program with actual grammar writers,
in comparison with a corpus-based method or machine learning.

As another line of investigation, we want to detect more kinds of conflicts.
For instance, say we have the following rules:

\begin{itemize}
\item[] \begin{verbatim}
REMOVE verb IF (-1C det) ;
REMOVE noun IF (-1C det) ;
\end{verbatim}
\end{itemize}

With our symbolic sentence, these rules will be no problem; to apply the latter, we only need to construct a target that has a realistic ambiguity with a noun; the verb is gone already.
However, in the context of real life texts, the grammar writer might actually mean the following:

\begin{itemize}
\item[] \begin{verbatim}
REMOVE verb IF (-1C det) (0 noun) ;
REMOVE noun IF (-1C det) (0 verb) ;
\end{verbatim}
\end{itemize}

In that case, those rules are contradictory, and we would like to find it out.

Finally, we want to test the approach to other grammar formalisms.

>>>>

We tested three grammars to find rules that cannot apply. 
The smallest grammar was
Dutch\footnote{\url{https://svn.code.sf.net/p/apertium/svn/languages/apertium-nld/apertium-nld.nld.rlx}},
with 59 rules; second was
Spanish\footnote{\url{https://svn.code.sf.net/p/apertium/svn/languages/apertium-spa/apertium-spa.spa.rlx}},
with 279 rules, and the largest was
Finnish\footnote{\url{https://github.com/flammie/omorfi/blob/master/src/vislcg3/omorfi.cg3}},
with 1185 rules. We left out \textsc{add}, \textsc{map} and other rule
types introduced in CG-3, and only tested the traditional
\textsc{remove} and \textsc{select} rules.
The results are shown in table~\ref{table:res}.

\begin{table}[]
\centering
\begin{tabular}{|l|l|l|l|}
\hline
                      & \textsc{nld}  & \textsc{spa}  & \textsc{fin}  \\ \hline
\# rules      & 59              & 279      & 1185     \\ \hline
\# readings      & 276             & 2191     & 2012    \\ \hline
\# conflicts       & 1               & 1        & \todo{n}    \\ \hline
\clock{} all rules       & 7s              & 3min 22s    & 3h 26min    \\ \hline
\clock{} last rule       & 4s              & 7s                & 3min 40s    \\ \hline
% This is with just 2000ish variables for Finnish, below is with 5700

@@

\paragraph{Correctness} 
The test revealed problems in all grammars, and for the smaller
grammars, we were able to verify manually that the detected rules were
true negatives. However, for the Finnish grammar, we did not have time
to investigate all of them; in addition, we suspect that some of
them are false negatives.
We did not check for false positives in any of the grammars.

@@

For the Finnish grammar, we could verify a number of tagset errors. 
The grammar was originally written in 1995, and converted to the
Omorfi tagset by \cite{pirinen2015}. For example, the Omorfi tagset
marks a possessive suffix with \texttt{Px}, e.g. \texttt{PxPl2} for second person plural, but the
grammar used the old convention \texttt{PL2} in the rules.
As a result, all rules with possessive suffixes were detected as
impossible.

Because of performance issues, we ignored the readings from the
lexicon and used only tag combinations that were named in the grammar.
For example, the rule \texttt{REMOVE (verb sg) IF (-1 det)}
gives us ``verb sg'' and ``det'', as opposed to fully specified
readings in the lexicon.
The shortcut will work most of the time, but it is possible to
miss some cases: e.g. \texttt{SELECT (verb sg) IF (0 p3)} 
requires a full ``verb sg p3'' reading, but our method only gives
``verb sg'' and ``p3'' separately. 
Some of the false negatives seem to come from this shortcut, or our naive
implementation of rules with \verb!*! in their context.
Due to these reasons, we cannot give conclusions about the Finnish
grammar; we include the results here only to have some preliminary
data on scaling up to larger grammars.
%For a condition situated $n$ or more words to the left or right, we only create a word exactly $n$ words away. This is something to be fixed in the future.

\paragraph{Running time} The speed of the check is an obvious issue,
especially for the Finnish grammar. 
As mentioned before, we took a shortcut and used 2000 readings from
the grammar.
With the full set of 6000 readings from the lexicon, the speed
was even worse: checking the last rule took 30 minutes, as opposed to
3 minutes with 2000 readings. 
%We never finished running the whole rule set with 6000 readings, but it would have taken potentially days.
However, the code is currently just a proof of concept, and we plan to
make an effort in optimisation.

@@

% The SAT problem for checking one rule grows depending on how many rules come before it, how long the context is, and how many tag combinations are there. The running time of the last rule gives some indication ...

% Not all questions are so expensive.
% For instance, checking the tag sets in the Finnish grammar doesn't require the context of the other rules at all,
%  The majority of the correctly detected errors in the Finnish grammar were rule-internal---these errors don't need the context of the other rules at all.

\paragraph{Future work}

We will work on the performance, and fix any errors we found in the
evaluation, such as the issue with non-exact position.
Another obvious improvement is to restrict possible ambiguities
in the symbolic words.
Currently we have only hard-coded some
very basic assumptions, such as ``word cannot be ambiguous between a
boundary and a real word'', but we want to extract these restrictions
automatically.
For instance, the Spanish lexicon contains the entries
\texttt{casa:casa<n><sg>}  and  \texttt{casa:casar<v><sg><p3>}, thus
we know that a word can be ambiguous between a noun and a verb.
By adding these restrictions to the symbolic words, we hope to
restrict the search space, and make the SAT problem easier to solve.

%We will also improve usability, by finding out which rules contrad
After this, we want to handle the full expressivity of CG-3, with \textsc{map}, \textsc{add} and \textsc{substitute} rules.
At that stage, it would be interesting to test our program with actual grammar writers,
in comparison with a corpus-based method or machine learning.
Finally, if the method proves feasible for Constraint Grammars, we want
to try applying it to other, more complex grammar formalisms.

% As another line of investigation, we want to detect more kinds of conflicts.
% For instance, say we have the following rules:

% \begin{itemize}
% \item[] \begin{verbatim}
% REMOVE verb IF (-1 det) ;
% REMOVE noun IF (-1 det) ;
% \end{verbatim}
% \end{itemize}

% With our symbolic sentence, these rules will be no problem; to apply the latter, we only need to construct a target that has a realistic ambiguity with a noun; the verb is gone already.
% However, in the context of real life texts, the grammar writer might actually mean the following:

% \begin{itemize}
% \item[] \begin{verbatim}
% REMOVE verb IF (-1 det) (0 noun) ;
% REMOVE noun IF (-1 det) (0 verb) ;
% \end{verbatim}
% \end{itemize}

% In that case, those rules are contradictory, and we would like to find it out.

### 2015-10-25 paper/lrec/lrec2016_listenmaa_claessen.tex
fonts, stuff
more cleanup

\documentclass[11pt]{article}
\usepackage{nodalida2015}
\usepackage{times}
\usepackage{mathptmx}
\usepackage{fixltx2e}
%\usepackage{txfonts}
\usepackage{url}
\usepackage{multirow}
\usepackage{latexsym}
\usepackage{cite}
%\usepackage{graphics}
\usepackage{graphicx}
\usepackage{slantsc}
\usepackage[colorlinks=true,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{authordate1-4}
\usepackage{multirow}
%\special{papersize=210mm,297mm} % to avoid having to use "-t a4" with dvips 
%\setlength\titlebox{6.5cm}  % You can expand the title box if you really have to

\usepackage{fontspec}

\newcommand{\mainfontfamily}{Palatino}
\setmainfont[
  Ligatures=TeX,
  Scale=0.95,
]{\mainfontfamily}
%\newfontfamily\scshape[Letters=SmallCaps, Numbers=Uppercase] % {Myriad Pro} %

\newcommand{\monofontfamily}{Monaco} % Courier
\setmonofont[Scale=0.8]{\monofontfamily}

\usepackage{color}
\newcommand{\todo}[1]{{\color{cyan}\textbf{[TODO: }#1\textbf{]}}}
% \usepackage{tikz}
% \def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) --
%   (.25,.15) -- cycle;}
% \newcommand{\Mypm}{\mathbin{\tikz [x=1.4ex,y=1.4ex,line width=.1ex] \draw (0.0,0) -- (1.0,0) (0.5,0.08) -- (0.5,0.92) (0.0,0.5) -- (1.0,0.5);}}%

\title{Analysing Constraint Grammars with a SAT solver}

@@

\begin{abstract}
We describe a method for
 analysing Constraint Grammars by encoding the
rules in SAT.
Our tools can detect internal conflicts or redundancies in a grammar,
as well as generate examples to demonstrate the effect of some rule or
rule set.
This can help users to diagnose and improve their grammars.
No corpus is required, only a morphological lexicon.
%We test our tools with grammars from 100-1000 rules.
\end{abstract}

>>>>

\documentclass[11pt]{article}
\usepackage{nodalida2015}
\usepackage{times}
\usepackage{mathptmx}
\usepackage{fixltx2e}
\usepackage{url}
\usepackage{multirow}
\usepackage{latexsym}
\usepackage{cite}
\usepackage{graphicx}
%\usepackage{slantsc}
\usepackage[colorlinks=true,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{authordate1-4}
\usepackage{multirow}
%\special{papersize=210mm,297mm} % to avoid having to use "-t a4" with dvips 
%\setlength\titlebox{6.5cm}  % You can expand the title box if you really have to

\usepackage{fontspec}
\usepackage{wasysym}

\newcommand{\mainfontfamily}{Palatino}
\setmainfont[
  Ligatures=TeX,
  Scale=0.9,
]{\mainfontfamily}

\newfontfamily\scshape[Letters=SmallCaps, Numbers=Uppercase]{Palatino
  Small Caps}

\newcommand{\monofontfamily}{Monaco} % Monaco
\setmonofont[Scale=0.8]{\monofontfamily}

\usepackage{color}
\newcommand{\todo}[1]{{\color{cyan}\textbf{[TODO: }#1\textbf{]}}}

\title{Analysing Constraint Grammars with a SAT-solver}

@@

\begin{abstract}
%\small
We describe a method for
 analysing Constraint Grammars by encoding the
rules in SAT.
Our tools can detect internal conflicts or redundancies in a grammar,
as well as generate examples to demonstrate the effect of some rule or
rule set.
This can help users to diagnose and improve their grammars.
No corpus is required, only a morphological lexicon.
%We test our tools with grammars from 100-1000 rules.
\end{abstract}

### 2015-10-25 paper/lrec/problem_desc.tex
more cleanup

As CGs grow larger, it gets harder for the grammar writers and users
to keep track of all the rules and their interaction.
With our tools, the users can diagnose and improve their grammars.
Some examples of conflicts are listed below:
\begin{itemize}
\item If a rule appears twice, the second occurrence will be disabled by the first
\item $R$ selects something in a context, $R'$ removes it
\item $R$ removes something from the context of $R'$, so $R'$ can never
  apply
\item $R$ has an internal conflict, such as non-existent
tag combination, or contradicting requirements for a context word
\end{itemize}
$R$ can also be a set of rules: for instance, if one rule removes a verb in
context $C$, and another in context $\neg C$, together these rules
remove a verb in all possible cases, disabling any future rule that
targets verbs.

%In addition to analysing the whole grammar,
Our method can be used to
test out individual rules. For instance, we can ask the tool to
generate example sequences that trigger rule(s) $R$ but not rule(s) $R'$. 
This functionality can help the grammar writer to see
if they have implemented the rules in the intended way.

>>>>

As CGs grow larger, it gets harder for the grammar writers and users
to keep track of all the rules and their interaction.
With our tools, the users can diagnose and improve their grammars.
Some examples of conflicts are listed below:
\begin{itemize}
\item If a rule appears twice, the second occurrence will be disabled by the first
\item $R$ selects something in a context, $R'$ removes~it
\item $R$ removes something from the context of $R'$, so $R'$ can never
  apply
\item $R$ has an internal conflict, such as non-existent
tag combination, or contradicting requirements for a context word
\end{itemize}
$R$ can also be a set of rules: for instance, if one rule removes a verb in
context $C$, and another in context $\neg C$, together these rules
remove a verb in all possible cases, disabling any future rule that
targets verbs.

%In addition to analysing the whole grammar,
Our method can be used to
test individual rules. For instance, we can ask the tool to
generate example sequences that trigger rule(s) $R$ but not rule(s) $R'$. 
This can help the grammar writer to see
if they have implemented the rules in the intended way.
